Efficient Thresholded Correlation using Truncated Singular Value Decomposition

### 2016-09-04 threshold/draft.Rnw
new arxiv copy; Hero reference; generic formatting
Anti-correlation section

%% Instructions for processing this document into a PDF document
%%
%% FIRST NOTE, you need the 'xtable' and 'biclust' packages installedi in R.
%% Some of the examples will come from data sets in this package.
%% You can install it from the R console with, for example:
%%
%% install.packages("biclust",repos="http://cran.case.edu")
%% install.packages("xtable",repos="http://cran.case.edu")
%%
%% Now you can prepare the PDF of the paper with any of:
%%
%% From a shell:
%% R CMD Sweave --pdf draft.Rnw
%%
%% From an R prompt:
%% Sweave("draft.Rnw")
%% tools::texi2pdf("draft.tex")
%%
%% From RStudio:
%% click on the little "Compile PDF" button
%%
%% Note that you should probably run the above a few times because
%% LaTeX generally takes a few runs to get reference numbers right.

@@

The number of correlation coefficients grows quadratically with the number of
vectors. Simply computing all pairs of correlation coefficients and then
filtering out coefficients below a given threshold may not be computationally
feasible for high-dimensional data in modern genomics and other applications.
Considerable attention has been devoted to \emph{pruning
methods} that cheaply prune away all but the most highly-correlated pairs of
vectors.  For example the TAPER and related algorithms of Xiong et al. develop
pruning rules based on the frequency of mutually co-occurring observations (rows)
among vector pairs; see \cite{prune1, prune2} and the references therein.
Related methods approximate a set containing the most highly-correlated
pairs using hashing algorithms \cite{prune3}. A number of other methods arrive
at approximate sets of the most highly-correlated pairs using
dimensionality-reduction techniques like the discrete Fourier
transform \cite{timeseries}. The method of Wu, et al. \cite{svd-similarity},
conceptually very similar to our approach, uses the singular value
decomposition (SVD) to find all pairs of close vectors with respect to a
distance metric (instead of correlation).  Our method finds all of the most
highly-correlated vector pairs with respect to a given threshold by pruning
along an intuitive path of decreasing variance defined by the data and computed
by a truncated SVD.

@@

\begin{figure}[!ht]
\begin{center}
<<eval=TRUE, echo=FALSE, fig=TRUE, height=5, width=7>>=
library(biclust)
data(EisenYeast)
A = t(EisenYeast)
A = sweep(A, 2, apply(A,2,mean), FUN=`-`)
A = sweep(A, 2, sqrt(apply(A,2,crossprod)), FUN=`/`)
s = svd(A)
i = order(s$v[,1])
plot(s$v[i,1], ylab=expression(P * v[1]), xlab="Sorted index", pch='.')
t=0.99
abline(h=c(0, sqrt(2*(1-t))/s$d[1]), col=2, lwd=1)
@
\caption{
Example plot of the ordered entries of $v_1$ for the EisenYeast example from
Section \ref{examples}. The red lines illustrate the interval corresponding
threshold $t=0.99$ placed at an arbitrary location on the vertical axis. Points
farther apart than this interval correspond to column vectors that do not meet
the correlation threshold.
\label{fig1}
}
\end{center}
\end{figure}

@@

\section{Extensions}\label{extensions}

>>>>

%% Instructions for processing this document into a PDF document
%%
%% FIRST NOTE, you need the 'xtable' and 'biclust' packages installed in R.
%% Some of the examples will come from data sets in this package.
%% You can install it from the R console with, for example:
%%
%% install.packages("biclust",repos="http://cran.case.edu")
%% install.packages("xtable",repos="http://cran.case.edu")
%%
%% Now you can prepare the PDF of the paper with any of:
%%
%% From a shell:
%% R CMD Sweave --pdf draft.Rnw
%%
%% From an R prompt:
%% Sweave("draft.Rnw")
%% tools::texi2pdf("draft.tex")
%%
%% From RStudio:
%% click on the little "Compile PDF" button
%%
%% Note that you should probably run the above a few times because
%% LaTeX generally takes a few runs to get reference numbers right.

@@

The number of correlation coefficients grows quadratically with the number of
vectors. Simply computing all pairs of correlation coefficients and then
filtering out coefficients below a given threshold may not be computationally
feasible for high-dimensional data in modern genomics and other applications.
Considerable attention has been devoted to \emph{pruning methods} that cheaply
prune away all but the most highly-correlated pairs of vectors.  For example
the TAPER and related algorithms of Xiong et al. develop pruning rules based on
the frequency of mutually co-occurring observations (rows) among vector pairs;
see \cite{prune1, prune2} and the references therein.  Related methods
approximate a set containing the most highly-correlated pairs using hashing
algorithms \cite{prune3}. A number of other methods arrive at approximate sets
of the most highly-correlated pairs using dimensionality-reduction techniques
like the discrete Fourier transform \cite{timeseries}. The method of Wu, et al.
\cite{svd-similarity}, conceptually very similar to our approach, uses the
singular value decomposition (SVD) to find all pairs of close vectors with
respect to a distance metric (instead of correlation). Hero and
Rajaratnam\cite{hero2011large} apply projections to the thresholded correlation
problem with the additional objective of determining a reasonable threshold
value; their method follows a probabilistic approach.  Our method finds all of
the most highly-correlated vector pairs with respect to a given threshold by
pruning along an intuitive path of decreasing variance defined by the data and
computed by a truncated SVD.

@@

\begin{figure}[!ht]
\begin{center}
<<eval=TRUE, echo=FALSE, fig=TRUE, height=5, width=7>>=
library(biclust)
data(EisenYeast)
A = t(EisenYeast)
A = sweep(A, 2, apply(A,2,mean), FUN=`-`)
A = sweep(A, 2, sqrt(apply(A,2,crossprod)), FUN=`/`)
s = svd(A)
i = order(s$v[,1])
plot(s$v[i,1], ylab=expression(P * v[1]), xlab="Permuted index", pch='.')
t=0.99
abline(h=c(0, sqrt(2*(1-t))/s$d[1]), col=2, lwd=1)
@
\caption{
Example plot of the ordered entries of $v_1$ for the EisenYeast example from
Section \ref{examples}. The red lines illustrate the interval corresponding
threshold $t=0.99$ placed at an arbitrary location on the vertical axis. Points
farther apart than this interval correspond to column vectors that do not meet
the correlation threshold.
\label{fig1}
}
\end{center}
\end{figure}

@@

\section{Practical extensions}\label{extensions}

This section presents a number of practical extensions to the main thresholded
correlation idea, most of which are available in the reference R package
implementation. Extensions include finding anti-correlated vectors, finding
thresholded correlations between column vectors from two different matrices,
and application to other distance metrics in place of correlation. All of the
discussed extensions are simple to implement.

\subsection{Anti-correlation}

Some applications are interested in finding both highly correlated and
anti-correlated vectors--that is, for a given threshold $t > 0$, find column
vectors $a_i$ and $a_j$ such that $|\mbox{cor}(a_i, a_j)| \ge t$.
Note that $-\mbox{cor}(a_i, a_j) = \mbox{cor}(a_i, -a_j)$, and that
similarly to Lemma \ref{sumlemma},
\begin{equation*}
\|a_i + a_j\|^2 =
\sigma_1^2 (e_i^Tv_{1} + e_j^Tv_{1})^2 + 
\sigma_2^2 (e_i^Tv_{2} + e_j^Tv_{2})^2 + \cdots + 
\sigma_k^2 (e_i^Tv_{k} + e_j^Tv_{k})^2,
\end{equation*}
where $e_j, v_j, \sigma_j$ are defined as in Lemma \ref{sumlemma}.
Then steps 2--4 of Algorithm \ref{prune} can be modified to include vector pairs meeting an
anti-correlation threshold as follows:
\begin{enumerate}
\item[2.] Let $\hat{v}_1\in\mathcal{R}^{2n}=[v_1; -v_1]$ (concatenate the vectors). Compute permutation
$P$ that orders points in $\hat{v}_1$.
\item[3.] Compute $\ell$, the longest run of successive points in $P \hat{v}_1$ within the interval $\sqrt{2(1-t)}/\sigma_1$. If $\ell > n-1$, then set $\ell = n-1$.
\item[4.] Compute a set of candidate index pairs $(\tilde{j},\tilde{j}+\tilde{i})$ with respect to the index order defined by the permutation $P$ that possibly meet the correlation threshold
\[
\bigcup_{\tilde{i}=1}^\ell
\left\{
(\tilde{j},\tilde{j}+\tilde{i}) :
\begin{array}{l}
e_{\tilde{j}}^T(D^{\tilde{i}} P V_{1:p} )^2 s_{1:p}^2 \le 2(1-t) \mbox{ or }
\\
e_{\tilde{j}}^T(|D^{\tilde{i}}| P V_{1:p} )^2 s_{1:p}^2 \le 2(1-t)
\end{array}
\right\},
\]
where $\tilde{j}=1,2,\ldots,n-\tilde{i}$, and $|D^{\tilde{i}}|$ indicates the entry-wise
absolute value of the matrix $D^{\tilde{i}}$.
\end{enumerate}
The modified step 3 produces a run length $\ell$ very likely to be larger than
in the default case, and step 4 above requires more arithmetic. Thus, including
anti-correlated vectors in the result is more expensive than simply finding
only highly correlated vectors.

\subsection{Thresholded correlation of columns between two different matrices}

WRITE ME

\subsection{Alternative metrics}

### 2016-09-04 threshold/draft.tex
new arxiv copy; Hero reference; generic formatting
Anti-correlation section

% latex table generated in R 3.2.2 by xtable 1.8-2 package
% Wed Mar  2 12:56:08 2016
\begin{table}[ht]
\centering
\begingroup\small
\begin{tabular}{lrr}
  \hline
Method & Wall clock time (s) & Peak excess memory use (MB) \\ 
  \hline
tcor & 3.13 & 59.60 \\ 
  brute force & 1.75 & 1075.10 \\ 
   \hline
\end{tabular}
\endgroup
\caption{Gene expression results, threshold=0.95} 
\label{EisenYeast}
\end{table}

@@

% latex table generated in R 3.2.2 by xtable 1.8-2 package
% Wed Mar  2 12:56:08 2016
\begin{table}[ht]
\centering
\begingroup\small
\begin{tabular}{lrrl}
  \hline
Method & Wall clock time (s) & Memory use (MB) &   \\ 
  \hline
tcor (4 core PC) & 10309.00 & 389.10 &  \\ 
  tcor (64 core cluster) & 482.00 & 994.30 & (local) \\ 
   &  & 42282.40 & (remote) \\ 
   \hline
\end{tabular}
\endgroup
\caption{TCGA ACC Methylation450 (80 rows x 394,014 columns, threshold=0.99). Linux cluster results include master R process memory use (local) plus total peak memory used by all 64 worker R processes (remote).} 
\label{TCGA3}
\end{table}From a set of almost 80 billion pairs of vectors, Algorithm \ref{prune} finds 913,601
pairs with correlation values of 0.99 or more on a modest Linux
cluster in about 8 minutes. The same algorithm works, more slowly, in
the constrained memory setting of a desktop PC without modification.

\section{Extensions}\label{extensions}

>>>>

% latex table generated in R 3.2.2 by xtable 1.8-2 package
% Sat Sep  3 20:12:12 2016
\begin{table}[ht]
\centering
\begingroup\small
\begin{tabular}{lrr}
  \hline
Method & Wall clock time (s) & Peak excess memory use (MB) \\ 
  \hline
tcor & 3.13 & 59.60 \\ 
  brute force & 1.75 & 1075.10 \\ 
   \hline
\end{tabular}
\endgroup
\caption{Gene expression results, threshold=0.95} 
\label{EisenYeast}
\end{table}

@@

% latex table generated in R 3.2.2 by xtable 1.8-2 package
% Sat Sep  3 20:12:12 2016
\begin{table}[ht]
\centering
\begingroup\small
\begin{tabular}{lrrl}
  \hline
Method & Wall clock time (s) & Memory use (MB) &   \\ 
  \hline
tcor (4 core PC) & 10309.00 & 389.10 &  \\ 
  tcor (64 core cluster) & 482.00 & 994.30 & (local) \\ 
   &  & 42282.40 & (remote) \\ 
   \hline
\end{tabular}
\endgroup
\caption{TCGA ACC Methylation450 (80 rows x 394,014 columns, threshold=0.99). Linux cluster results include master R process memory use (local) plus total peak memory used by all 64 worker R processes (remote).} 
\label{TCGA3}
\end{table}From a set of almost 80 billion pairs of vectors, Algorithm \ref{prune} finds 913,601
pairs with correlation values of 0.99 or more on a modest Linux
cluster in about 8 minutes. The same algorithm works, more slowly, in
the constrained memory setting of a desktop PC without modification.

\section{Practical extensions}\label{extensions}

This section presents a number of practical extensions to the main thresholded
correlation idea, most of which are available in the reference R package
implementation. Extensions include finding anti-correlated vectors, finding
thresholded correlations between column vectors from two different matrices,
and application to other distance metrics in place of correlation. All of the
discussed extensions are simple to implement.

\subsection{Anti-correlation}

Some applications are interested in finding both highly correlated and
anti-correlated vectors--that is, for a given threshold $t > 0$, find column
vectors $a_i$ and $a_j$ such that $|\mbox{cor}(a_i, a_j)| \ge t$.
Note that $-\mbox{cor}(a_i, a_j) = \mbox{cor}(a_i, -a_j)$, and that
similarly to Lemma \ref{sumlemma},
\begin{equation*}
\|a_i + a_j\|^2 =
\sigma_1^2 (e_i^Tv_{1} + e_j^Tv_{1})^2 + 
\sigma_2^2 (e_i^Tv_{2} + e_j^Tv_{2})^2 + \cdots + 
\sigma_k^2 (e_i^Tv_{k} + e_j^Tv_{k})^2,
\end{equation*}
where $e_j, v_j, \sigma_j$ are defined as in Lemma \ref{sumlemma}.
Then steps 2--4 of Algorithm \ref{prune} can be modified to include vector pairs meeting an
anti-correlation threshold as follows:
\begin{enumerate}
\item[2.] Let $\hat{v}_1\in\mathcal{R}^{2n}=[v_1; -v_1]$ (concatenate the vectors). Compute permutation
$P$ that orders points in $\hat{v}_1$.
\item[3.] Compute $\ell$, the longest run of successive points in $P \hat{v}_1$ within the interval $\sqrt{2(1-t)}/\sigma_1$. If $\ell > n-1$, then set $\ell = n-1$.
\item[4.] Compute a set of candidate index pairs $(\tilde{j},\tilde{j}+\tilde{i})$ with respect to the index order defined by the permutation $P$ that possibly meet the correlation threshold
\[
\bigcup_{\tilde{i}=1}^\ell
\left\{
(\tilde{j},\tilde{j}+\tilde{i}) :
\begin{array}{l}
e_{\tilde{j}}^T(D^{\tilde{i}} P V_{1:p} )^2 s_{1:p}^2 \le 2(1-t) \mbox{ or }
\\
e_{\tilde{j}}^T(|D^{\tilde{i}}| P V_{1:p} )^2 s_{1:p}^2 \le 2(1-t)
\end{array}
\right\},
\]
where $\tilde{j}=1,2,\ldots,n-\tilde{i}$, and $|D^{\tilde{i}}|$ indicates the entry-wise
absolute value of the matrix $D^{\tilde{i}}$.
\end{enumerate}
The modified step 3 produces a run length $\ell$ very likely to be larger than
in the default case, and step 4 above requires more arithmetic. Thus, including
anti-correlated vectors in the result is more expensive than simply finding
only highly correlated vectors.

\subsection{Thresholded correlation of columns between two different matrices}

Arbitrary vectors

\subsection{Alternative metrics}
